---
title: "Cell Differentiation"
author: "Gianluca"
date: "2025-03-21"
output: html_document
---

```{r setup}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(rtracklayer)
})

setwd("C:/Users/gianl/Desktop/ETH/BiologyMsc/Spring Semester Material/BARGE/Assignments/week5")
```

# Differentiated Cells
Since we are using only H3K4me3 and H3K27me3 regions I will consider only TSS bivalent domains, and not also enhancers which are marked by H3K4me1 and H3K27me3 histone modifications.
```{r download}
# Mouse Embryonic fibroblast H3K27me3
download.file("https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz", "ENCFF008XKX.bed.gz")

# Mouse Embryonic fibroblast H3K4me3
download.file("https://www.encodeproject.org/files/ENCFF357JNZ/@@download/ENCFF357JNZ.bed.gz", "ENCFF357JNZ.bed.gz")
```

# Extracting peaks from bed files
question how can I use lapply in this case? I should rename all bed files and include them in a folder that I can refer to using peakfiles <- list.files("peaks", full.names = TRUE) to create the list for lapply?

```{r peaks}
#ES H3K4me3
peaks1ES <- rtracklayer::import("ENCFF974BMC.bed.gz", format="narrowPeak")

# ES H3K27me3
peaks2ES <- rtracklayer::import("ENCFF274UIB.bed.gz",format="narrowPeak")

# Fib H3K4me3
peaks1Fib <- rtracklayer::import( "ENCFF357JNZ.bed.gz", format="narrowPeak" )

# Fib H3K27me3
peaks2Fib <- rtracklayer::import("ENCFF008XKX.bed.gz", format="narrowPeak")
```

# Find bivalent domains (BD)
The idea is to find regions in ES cells where both markers overlap and save the overall genetic locus in a list that can then be used again to identify new overlapping regions with the same markers on Fibroblasts.
```{r bivalent domains}
overlaps_ES <- findOverlapPairs(peaks1ES, peaks2ES)
all_ranges <- c(first(overlaps_ES), second(overlaps_ES))
ESBD_TSS <- reduce(all_ranges)

FibOverlaplist <- findOverlapPairs(peaks1Fib, peaks2Fib)
all_ranges2 <- c(first(FibOverlaplist), second(FibOverlaplist))
FibBD_TSS <- reduce(all_ranges2)
```

# Amount of overlapping BDs in differentiated cells
```{r}
overlapping_BD <- sum(overlapsAny(ESBD_TSS, FibBD_TSS))
overlapping_H3K4me3 <- sum(overlapsAny(peaks1Fib, ESBD_TSS))
overlapping_H3K27me3 <- sum(overlapsAny(peaks2Fib, ESBD_TSS))

overlapping_BD
overlapping_H3K4me3
overlapping_H3K27me3
```

# Report

Mouse embryonic stem cell bivalent domains overlap with H3K4me3&H3K27me3 domains in only 44 cases. In contrast, ES bivalent domains overlapping only H3K4me3 are 9395, while those overlapping only H3K27me3 are 45.