---
title: "assignment"
author: "Gianluca"
date: "2025-02-28"
output: html_document
---
# Packages required
```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(GenomicRanges)
  library(ensembldb)
  library(epiwraps)
  library(ggplot2)
})
```

# Downloading M musculus files
Finding M musculus EnsDb object, version 102, saving the ID and downloading the annotation file
```{r}
ah <- AnnotationHub()
query_results <- query(ah, c("GRCm38","EnsDb"))
query_results
# version AH89211
mmEns <- ah[["AH89211"]]
```

Downloading M musculus genome sequence in TwoBit format
```{r}
qr2 <- query(ah, c("GRCm38","dna_sm"))
qr2
# version AH88477
mmGenome <- ah[["AH88477"]]
```
# Dowloading Drosophila melanogaster genome
```{r}
qr3 <- query(ah, c("BDGP6", "dna_sm"))
qr3
# version AH106160
dmGenome <- ah[["AH106160"]]
```
# M musculus assignments
## Counting gene IDs and symbols of PCGs
```{r}
PCGs <- genes(mmEns, filter=GeneBiotypeFilter("protein_coding"))
head(PCGs)
length(unique(PCGs$gene_id))
# Gene ID count: 22287
length(unique(PCGs$symbol))
# Symbols count 21964
```
## Distribution of spliced lengths
```{r}
ex <- exonsBy(mmEns, by ="tx", filter = GeneBiotypeFilter("protein_coding"))
ex_length <- sum(width(ex))
head(ex_length)


lengths_df <- data.frame(length = ex_length)
ggplot(lengths_df, aes(x=length))+
  geom_histogram(binwidth = 100, fill = "red")+
  labs(title="Distribution of Spliced Length of PCGs")+
  xlab("Spliced Length of PCGs")+
  ylab("Count of Transcripts")
```
By looking at the sorted dataframe, there are few transcripts with a length of > 40000 nucleotides.
```{r}
count(lengths_df$length > 20000)
# Only 36 transcripts are longer than 20000, so I want to plot everything thats under 20000

lengths_red <- lengths_df[lengths_df$length<20000,]
length_df_red <- data.frame(length = lengths_red)

ggplot(length_df_red, aes(x=length))+
  geom_histogram(binwidth = 100, fill = "red")+
  labs(title="Distribution of Spliced Length of PCGs")+
  xlab("Spliced Length of PCGs")+
  ylab("Count of Transcripts")
```


